---
title: "Geog4/6300: Lab 1"
format: gfm
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

## Loading data into R, data transformation, and summary statistics

**Mary Elizabeth Rudd**

**Overview and lab criteria:**

This lab is intended to assess your ability to use R to load data and to generate basic descriptive statistics. For this lab to be marked complete, the following criteria must be met:

4. Identify and apply appropriate data filtering and cleaning strategies to prepare datasets for analysis. (Task 2)
5. Effectively interpret the code you create, explaining in plain language what each step does to the data. (Task 6)
6. Identify and use appropriate external documentation — including package references, help files, and peer resources — to learn and apply unfamiliar functions or methods. (Task 7)
7. Filter, aggregate, and transform datasets using grouping and summary operations to answer specific analytical questions. (Tasks 1 & 5)
9. Reshape data between wide and long formats to meet the requirements of different analytical or visualization tasks. (Task 4)
11. Create effective data visualizations across multiple chart types (line, scatter, histogram, Q-Q plot), applying appropriate aesthetic choices such as color, grouping, and labeling. (Task 3 & 5)

**Data:**

You'll be using monthly weather data from the Daymet climate database (http://daymet.ornl.gov) for all counties in the United States over an 12-year period (2010-2021). These data are available on the GitHub repo for our course. The following variables are provided:

* `cty_txt`: Code for joining to census data
* `year`: Year of observation (with an initial "Y" to make it a character)
* `month`: Month of observation (1 = Jan, 2 = Feb, etc.)
* `median_tmax`: Median maximum recorded temperature (Celsius)
* `median_tmin`: Median minimum recorded temperature (Celsius)
* `sum_prcp`: Total recorded precipitation for the month (mm)
* `cty_name`: Name of the county
* `state`: State of the county
* `region`: Census region (map: https://www2.census.gov/geo/pdfs/maps-data/maps/reference/us_regdiv.pdf)
* `division`: Census division
* `X`: Longitude of the county centroid
* `Y`: Latitude of the county centroid

These labs are meant to be done collaboratively, but your final submission should demonstrate your own original thought (don't just copy your classmate's work or turn in identical assignments). Your answers to the lab questions should be typed in this Quarto template. You'll then render the document to a GitHub markdown document and upload it to your class GitHub repo.

**Procedure:**

Load the tidyverse package and import the data:

```{r}
#| message: false

library(tidyverse)

daymet_data <- read_csv("data/daymet_monthly_median_2010-2021.csv")
```

We can look at the first few rows of the dataset using the *head()* function. We also use *kable* to format the output as a readable table..

```{r}
kable(head(daymet_data))
```

There are a lot of observations here, 452,448 to be exact. To get a better grasp on the data, we can use `group_by()` and `summarise()` from the tidyverse package. This will allow us to identify the mean value for each year by county across the study period.

## Task 1

*Use `group_by()` and `summarise()` to calculate the mean minimum temperature for each year by county across all months, also including State and Region as grouping variables. Your resulting dataset should show the value of tmin for each county in each year. Use the `kable()` and `head()` functions as shown above to call the resulting table.*

```{r}
#| label: task1

daymet_tmin <- daymet_data %>%
  group_by(state, region, cty_name, year) %>%
  summarise(mean_tmin = mean(median_tmin))

kable(head(daymet_tmin))

```

## Task 2

*Let's shift to the state level, focusing on those in the South Region. Filter the original data frame (`daymet_data`) to just include counties in this region. Then calculate the mean minimum temperature by year for each state. For an optional extra challenge, use the `round()` function to include only 1 decimal point. Use `kable()` and `head()` to call the first few lines of the resulting table.*

```{r}
#| label: task2

daymet_south <- daymet_data %>%
  filter(region == "South Region") %>%
  group_by(state,year) %>%
  summarise(mean_tmin = round(mean(median_tmin), 1))

kable(head(daymet_south))
```

## Task 3

*To visualize the trends, we could use ggplot to visualize change in mean temperature over time. Create a line plot (`geom_line`) showing the state means you calculated in task 2. Use the `color` parameter to show separate colors for each state. You may also need to define the state as a group in the aesthetic parameter.*

```{r}
#| label: task3

  ggplot(daymet_south, 
         aes(x=year,
             y=mean_tmin,
             color=state,
             group=state))+
  geom_line() +
  labs(title="Mean Minimum Temperature by State Over Time",
       x="Year",
       y="Mean Minimum Temp (Cº)") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Task 4

*If you wanted to look at these data as a table, you'd need to have it in wide format. Use the `pivot_wider()` function to create a wide-format version of the data frame you created in task 2. In this case, the rows should be states, the columns should be the years, and the data in those columns should be mean minimum temperatures. Then call the whole table using `kable()`.*

```{r}
#| label: task4

daymet_south_wide<-daymet_south %>%
  pivot_wider(id_cols="state",
               names_from="year",
               values_from="mean_tmin")

kable(daymet_south_wide)
```

## Task 5

*Let's assess the relationship of heat and precipitation by region. Returning to the original dataset, create a data frame that shows the mean maximum temperature and mean precipitation for all states in 2015, also including region as a subgroup in your `group_by`. Then use ggplot to create a scatterplot (`geom_point`) for these two variables, coloring the points using the region variable.*

```{r}
#| label: task5

daymet_regions <- daymet_data %>%
  filter(year=="Y2015")%>%
  group_by(state,year,region) %>%
  summarise(mean_tmax = round(mean(median_tmax), 1),
           mean_prcp= round(mean(sum_prcp), 1))

ggplot(daymet_regions, 
         aes(x=mean_tmax,
             y=mean_prcp,
             color=region))+
  geom_point() +
  labs(title="Mean Maximum Temperature and Mean\nMonthly Precipitation Amount by Region in 2015",
       x="Mean Maximum Temperature (Cº)",
       y="Mean Monthly Precipitation Amount (mm)") + 
  theme(plot.title = element_text(hjust = 0.5, size = 14))
        
  
```

## Task 6

*In the space below, explain what each function in your code for task 5 does to the dataset in plain English.*

1. daymet_regions <- daymet_data %>% - this takes all the functions done below and applies them to a new daymet_regions dataframe
2. filter() - filters out all other years except Y2015 from the data
3. group_by() - splits the data into groups by state/region/year
4. summarise() turns each of the groups into one row and calculates and mean round inside of those rows in the tmax and pcrp columns
5. aes () links columns to visual properties like axes, colors, etc...
6. geom_point() draws the dots/places the data on the plot
7. labs() allows you to change labels, axis titles, legends
8. theme() allows you to change appearance of plot elements


First section:My first section of code filters the daymet_data dataframe to isolate datapoints from the year 2015 and creates a new dataframe called daymet_regions. I then grouped this filtered data by state, year, and region to make sure the data is grouped by those variables, and then used a function (summarise) that takes the mean of the median_tmax and sum_prcp fields and rounds those values to 1 digit right of the decimal point.

Second section: In this section, I created a scatter plot that displayed each state's mean maximum temperature (x-axis) and their mean precipitation amount (y-axis) for 2015. I added color so that the dots on the scatterplot are colored corresponding to their regions, i.e. the state names aren't visible anywhere. This creates a scatterplot that easily shows the patterns in average precipitation and temperatures in each region. I added a plot title with a manual line break and titles for my x and y axes, as well as centered and set a size for my plot title.

## Task 7

*The `dplyr` package also includes `across` function. Use `?across` on the R command line to open the documentation for this function. In the space below, explain what it does in your own words. Then interpret the way the across function is used below, going line by line within the function.*

```{r}
state_2015 <- daymet_data %>%
  filter(year == "Y2015") %>%
  group_by(region, state) %>%
  summarise(
    across(
      c(median_tmax, sum_prcp),
      mean,
      na.rm = TRUE,
      .names = "mean_{.col}"
    )
  )
```

The 'across' function allows you to apply one transformation to multiple columns at the same time. In the function above:
Line 180 calls the function within the "summarise" function, meaning we are wanting to summarize multiple columns at the same time/not having to call a function for each singular column we want to work with
Line 181 selects the columns median_tmax and sum_prcp 
Line 182 calculates the mean of median_tmax and sum_prcp at the same time and within one singular function
Line 183 excludes missing (NA) values
Line 184 is setting a naming convention for the columns that summarise(across)) is being applied to. It inserts "mean_" in front of every column name, creating mean_median_tmax and mean_sum_prcp


## Challenge Question

In class, we covered ways of working with the Daymet API. Create a script below that uses the **daymetr** package to download data from Daymet for a place (or places) of your choosing. Then visualize the temporal pattern for a variable of your choosing in this place, similar to what you did in question 4. Use a dplyr function (`mutate()`, `summarise()`, `filter()`, etc.) to do any needed data wrangling and create a visual using ggplot.

In addition to this code, write a short summary of a pattern that's evident in the data you visualized.

```{r}
#| label: challenge

```

*Explanation goes here.*

## Final Submission Stuff

### Disclosure of Assistance

Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?

I used Claude for question 2 because I couldn't figure out where to put the round function, and that's it!

### Lab Reflection

How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What are the biggest things you learned by completing it?

This lab was honestly a little bit easier than the previous lab, simply because I am much more familiar with RStudio now after having more practice with it. On a scale of 1-10 this was probably a 3. 